---
title: "EPS 704 Week 1: Data Import and Entry"
author: "Miguel Hernandez"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---
**GitHub repository:**
https://github.com/mah1201-png/EPS704.git
<!-- BEFORE YOU KNIT:
     1. Put this .Rmd file and the data file in the SAME folder.
     2. In RStudio click: Session > Set Working Directory > To Source File Location.
     3. Click the Knit button to build the PDF and the HTML. -->

```{r setup, include=FALSE}
# include=FALSE hides this setup chunk from the final report.
knitr::opts_chunk$set(echo = TRUE)   # echo = TRUE prints your code in the report.
```

## Goal

This report imports two data files and confirms that each one loaded correctly.

## Import a CSV file

```{r import-csv}
library(readr)                          # readr holds read_csv.
enroll <- read_csv("enrollment.csv")    # read the file into a table named enroll.
head(enroll)                            # show the first rows in the report.
dim(enroll)                             # show the row count and the column count.
```

The enrollment file loaded with `r nrow(enroll)` rows and `r ncol(enroll)` columns.

## Import one Excel sheet

```{r import-excel}
library(readxl)                         # readxl holds read_excel.
demo <- read_excel("institutional.xlsx",# read the workbook.
                   sheet = "Demographics") # name the sheet to read.
head(demo)                              # show the first rows in the report.
dim(demo)                               # show the row count and the column count.
```

The Demographics sheet loaded with `r nrow(demo)` rows and `r ncol(demo)` columns.

## Note

The text above uses inline code such as `r nrow(enroll)` so the numbers update
themselves every time you knit. Click Knit to build the PDF and the HTML.
