---
title: "EPS 704 Week 4: Descriptive Statistics"
author: "Miguel Hernandez"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

<!-- BEFORE YOU KNIT: put this file and the data in the same folder, then in RStudio
     click Session > Set Working Directory > To Source File Location, then click Knit. -->

**GitHub repository:** https://github.com/mah1201-png/EPS704

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)   # echo = TRUE prints your code in the report.
```

## Goal

Describe gpa, count categories, build a two way table, and a correlation matrix.

## Summaries and tables

```{r describe}
library(readr); library(dplyr)
enroll <- read_csv("enrollment.csv")
enroll <- enroll |> mutate(gpa = na_if(gpa, 999))    # 999 is a sentinel for missing gpa.

summary(enroll$gpa)                                  # five number summary.
sd(enroll$gpa, na.rm = TRUE)                         # spread of gpa.
table(enroll$program)                                # students per program.
table(enroll$program, enroll$degree_level)           # two way table.

enroll |> group_by(program) |>
  summarise(mean_gpa = mean(gpa, na.rm = TRUE), students = n())   # average gpa per program.

cor(enroll[c("gpa","credits","gre_q","gre_v")], use = "complete.obs")  # matrix.
```

The average gpa is `r round(mean(enroll$gpa, na.rm = TRUE), 2)`.

## Bonus: strongest and weakest correlation

The strongest link in the correlation matrix is between `gre_q` and `gre_v` (r approximately 0.04), and the weakest is between `credits` and `gre_v` (r approximately -0.006). Overall, none of these variables show a meaningful linear relationship with each other — all correlations are close to zero.