---
title: "EPS 704 Week 5: Exploratory Analysis and Statistics"
author: "Miguel Hernandez"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

<!-- BEFORE YOU KNIT: put this file and the data in the same folder, then in RStudio
     click Session > Set Working Directory > To Source File Location, then click Knit. -->

**GitHub repository:** https://github.com/mah1201-png/EPS704

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)   # echo = TRUE prints your code in the report.
```

## Goal

Run correlation, a t test, ANOVA, chi square, regression, and a paired test.

## Tests

```{r tests}
library(readr)
library(dplyr)
enroll <- read_csv("enrollment.csv")
enroll <- enroll |> mutate(gpa = na_if(gpa, 999))    # 999 is a sentinel for missing gpa.

cor.test(enroll$gpa, enroll$credits)                 # correlation with p value.
t.test(gpa ~ graduated, data = enroll)               # two groups.
enroll |> group_by(graduated) |> summarise(mean_gpa = mean(gpa, na.rm = TRUE))   # bonus: group means.
summary(aov(gpa ~ program, data = enroll))           # many groups.
chisq.test(table(enroll$degree_level, enroll$graduated))   # two categories.
summary(lm(gpa ~ credits, data = enroll))            # regression.

pp <- read_csv("pre_post_scores.csv")
t.test(pp$post, pp$pre, paired = TRUE)               # paired test.
```

The correlation between gpa and credits is weak and not statistically significant, r is about -0.04 and p is 0.26.

The t test shows no significant difference in gpa between graduated and non-graduated students, p is 0.71, with average gpa of about 3.33 in both groups.

The ANOVA shows a statistically significant difference in gpa across programs, p is less than 0.0001.

The chi square test shows no significant link between degree level and graduation status, p is 0.86.

The regression shows credits does not meaningfully predict gpa, the slope is close to zero and p is 0.26.

The paired t test shows a significant increase from pre to post scores, p is less than 0.0001, with an average increase of about 4.18 points.